Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance

### 2017-07-27 ms_pre_post.Rmd
Edit through Results

After costs are realised, each remaining female chooses $N_{males}$ males to mate with, where $N_{males}$ is calculated by sampling from a Poisson distribution such that $N_{males} = Poisson(P_{p}) + 1$. This ensures that all surviving females choose at least one mate and generates each female's realised degree of polyandry with some stochastic variation around the expected mean $N_{males}$ of $P_{p}+1$.

@@

Most often, $N_{males}$ will be smaller than the total number of available males [@Duthie]. Each female then chooses her $N_{males}$ mates based on her pre-copulatory inbreeding avoidance phenotype ($M_{p}$). Negative or positive $M_{p}$ values cause a female to avoid or prefer mating with kin, respectively, whereas $M_{p}=0$ cause females to mate randomly with respect to kinship.

@@

The probability that an offspring of female $i$ is sired by any one of $i$'s mates $j$ is calculated by assigning a perceived fertilisation quality to each $j$ with respect to $i$, $Q^{f}_{ij}$. Perceived fertilisation quality $Q^{f}_{ij}$ is calculated in the same way as perceived mate quality $Q^{m}_{ij}$, such that if female $i$ has a strategy of post-copulatory inbreeding avoidance ($F_{p}<0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = (-F_{p} \times k_{ij} + 1)^{-1}$. If the female has strategy of post-copulatory inbreeding preference ($F_{p}>0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = F_{p} \times k_{ij} + 1$. A relative quality ($q^{f}_{ij}$) is then calculated for each male by dividing his $Q^{f}_{ij}$ by the sum of the $Q^{f}_{ij}$ values across all of a female's mates. These $q^{f}_{ij}$ values, which lie between zero and one, define the probability of paternity. Females produce $n$ offspring, so a female $i$ samples from her mates $n$ times independently and with replacement with a probability of $q^{f}_{ij}$ for each male $j$ to determine the realised distribution of sires. Offspring have equal probability of being female or male. After offspring production, all female and male adults die so that generations are non-overlapping.

@@

The viability of a female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a log-linear function of her kinship with the sire $j$ of the offspring ($k_{ij}$) and inbreeding depression slope $\beta$,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
Here, $\beta$ models the number of haploid lethal equivalents that exist as deleterious recessive alleles in the gametes of $i$ and $j$, and which might be homozygous in offspring and reduce viability. Equation 1 assumes independent allelic effects, generating multiplicative effects on offspring viability [@Morton1956; @Mills1994]. It also assumes that inbreeding does not covary with inbreeding load (i.e., no purging), allowing the effects of kinship on inbreeding depression to be consistent across replicate simulations. Additionally, previous modelling has found that inbreeding avoidance in biparental populations has a negligible effect on load [@Duthie2016a]. We model inbreeding depression as having an absolute rather than relative effect on offspring viability (i.e., hard rather than soft selection), so that the effect of $\beta$ is consistent across generations and different parameter combinations. We assume that inbreeding always decreases offspring viability (i.e., $\beta > 0$, giving inbreeding depression but no outbreeding depression). Therefore, because $0 \leq k_{ij} \leq 1$, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero (if $-\beta \times k_{ij}$ is very negative) and one (if $-\beta \times k_{ij} = 0$). We therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability (versus mortality) using a Bernoulli trial. Offspring that are viable after the Bernoulli trial become adults. Given our current objectives, we simulate evolution under conditions where inbreeding avoidance is adaptive due to strong inbreeding depression, and not evolution where inbreeding preference is adaptive due to weak or zero inbreeding depression [@Parker1979; @Kokko2006; @Duthie2016a], or given outbreeding depression [@Bateson1983; @Greeff2009]. However, as described above, positive $M_{p}$ and $F_{p}$ values resulting in inbreeding preference are not precluded from evolving, and could potentially arise due to mutation or drift.

@@

Simulations and analysis
----------------------------------------------

@@

To address how asymmetric costs associated with pre-copulatory and post-copulatory inbreeding avoidance and polyandry affect the long-term persistence of reproductive strategies, we quantify the change in $M_{a}$ and $F_{a}$ over generations in simulations where pre-copulatory inbreeding strategy was cost free ($c_{M}=0$) but post-copulatory inbreeding strategy was moderately costly ($c_{F}=0.02$), and vice versa. We compare evolutionary trajectories with those of a costly strategy in the absence of evolution of an alternative strategy (e.g., evolution of pre-copulatory inbreeding strategy when post-copulatory inbreeding strategy phenotype is fixed at zero, $F_{p}=0$). Previous modelling using similar genetic architecture suggests that a cost of $0.02$ imposes strong but not overwhelming direct negative selection on polyandry [@Duthie]. This value is therefore appropriate to illustrate the different evolutionary consequences that could result from asymmetrical costs (but see Supporting Information for other asymmetric cost combinations).

To address how selection on an initially rare strategy of pre-copulatory or post-copulatory inbreeding avoidance, and resulting evolution, is affected by the other strategy of inbreeding avoidance already being fixed in the population, we first used exploratory simulations to quantify evolution of pre-copulatory inbreeding strategy, and of post-copulatory inbreeding strategy and associated polyandry, in isolation. Then, to test whether pre-copulatory inbreeding avoidance would evolve when adaptive polyandry and post-copulatory inbreeding avoidance were fixed, we initiated $M_{a}$ allele values at zero, but fixed $F_{a}$ allele values at $-10$ and $P_{a}$ allele values at $1$ (i.e., $F_{p}$ and $P_{p}$ were expressed did not evolve further). Similarly, to test whether post-copulatory inbreeding avoidance would evolve when pre-copulatory inbreeding avoidance was already fixed, we initiated $F_{a}$ and $P_{a}$ allele values at zero but fixed $M_{a}$ allele values at $-10$. Consequently, because females have 10 diploid loci, when $M_{a}$ or $F_{a}$ alleles were fixed at $-10$, outbred females were $51$ times less likely to choose a full brother and $13.5$ times times less likely to choose a first cousin than a non-relative in pre-copulatory and post-copulatory choice, respectively.

In all simulations, we recorded mean values of $P_{a}$, $M_{a}$, and $F_{a}$ in each generation and present these values over generations to infer selection on phenotypes ($P_{p}$, $M_{p}$, and $F_{p}$). Each combination of parameter values simulated was replicated $40$ times, and grand mean values and standard errors of means are calculated in each generation across replicates. These analyses allowed us to infer how allele values changed over generations in response to costs, but also in response to the changing values of other alleles and therefore potential evolutionary feedbacks among reproductive strategies. In our analyses, we do not use statistical tests to interpret simulation results; such tests are innappropriate because they violate the underlying assumptions of statistical hypothesis testing, and because statistical power (and therefore p-values) would be entirely determined by the number of simulation replicates performed [@White2014]. For all replicates, we set the maximum number of generations to 40000, which exploratory simulations and previous modelling [@Duthie] showed to be sufficient for inferring long-term dynamics of mean allele values and therefore selection on phenotypes.  For all replicates, we set $\rho=5$ immigrants, which produced a range of kin and non-kin in each generation allowing females to express inbreeding strategies, and $n=8$ offspring, which was sufficient to keep populations consistently at carrying capacities and avoid population extinction. Values of $K_{f}$ and $K_{m}$ were both set to 100 because previous modelling showed that populations of this size are small enough that mate encounters between kin occur with sufficient frequency for selection on inbreeding strategy, but not so small that selection is typically overwhelmed by drift [@Duthie2016a].

@@

However, when both pre-copulatory and post-copulatory inbreeding avoidance could evolve, their relative evolutionary dynamics depended on their relative costs. When pre-copulatory but not post-copulatory inbreeding avoidance was costly ($c_{M}=0.02$ and $c_{F}=0$), pre-copulatory inbreeding avoidance initially evolved (i.e., $M_{a} < 0$) but then evolved back towards random mating (i.e., $M_{a} \approx 0$) following increasing evolution of post-copulatory inbreeding avoidance and polyandry (Figure 3B). Similarly, when post-copulatory but not pre-copulatory inbreeding avoidance was costly ($c_{F}=0.02$ and $c_{M}=0$), post-copulatory inbreeding avoidance initially evolved (i.e., $F_{a} < 0$) before evolving back to random fertilisation (i.e., $F_{a} \approx 0$) after ca $20000$ generations (Figure 3D). This evolution of a high cost strategy back toward random mating or fertilisation in the presence of a relatively low cost alternative inbreeding strategy is a general consequence of differential selection on each strategy induced by cost asymmetry. Hence, this pattern can be observed for both non-zero low relative costs (e.g., values of 0.01 and 0.03; Figure S9) and smaller cost asymmetries (e.g., values of 0.01 and 0.02; Figure S10).

When pre-copulatory inbreeding avoidance was costly, allowing evolution of cost free post-copulatory inbreeding avoidance greatly facilitated evolution of polyandry (Figure 3A versus 3B). However when post-copulatory inbreeding avoidance was costly, allowing evolution of cost free pre-copulatory inbreeding avoidance caused $P_{a}$ alleles to decrease to very negative values, reducing expression of polyandry (Figure 3C versus 3D; polyandry was cost free in all these simulations). Results for all possible cost combinations of $0$ and $0.02$ are provided in Supporting Information (Figure S11).

@@

When polyandry alleles ($P_{a}$) were fixed to be positive so that all females were expected to mate multiply and post-copulatory inbreeding allele ($F_{a}$) values were fixed for adaptive inbreeding avoidance, pre-copulatory inbreeding avoidance evolved (i.e., $M_{a}$ values became increasingly negative; Figure 4A). Such evolution still occurred, but to a much smaller degree, when pre-copulatory inbreeding avoidance was costly (Figure 4B). However, after $40000$ generations, $M_{a}$ values were less negative when post-copulatory inbreeding avoidance and polyandry were fixed at non-zero values than when they also evolved from initial values of zero ($-15.45$ vs. $-17.62$; compare the blue lines in Figures 4A versus Figure 1B). Hence, we hypothesise that existence of fixed post-copulatory inbreeding avoidance can weaken selection and subsequent evolution of pre-copulatory inbreeding avoidance.

Conversely, when pre-copulatory inbreeding allele ($M_{a}$) values were fixed for adaptive inbreeding avoidance, mean $F_{a}$ allele values did not consistently become negative over generations (Figure 4C,D). Hence, we hypothesise that existence of fixed pre-copulatory inbreeding avoidance can prevent evolution of post-copulatory inbreeding avoidance. In our model, mean $P_{a}$ allele values consistently decreased over generations, reflecting selection against polyandry regardless of whether or not post-copulatory inbreeding avoidance was costly (Figure 4C,D). Consequently, when $c_{F}=0$, $F_{a}$ allele values had no effect because females were almost exclusively monandrous, resulting in high drift of $F_{a}$ values (resulting in variation among replicates illustrated by the wide standard errors in Figure 4C). However, when $c_{F}=0.02$, $F_{a}$ values remained near zero to minimise direct costs. This lack of selection for post-copulatory inbreeding avoidance was driven by a lack of polyandry, and therefore an inability of females to bias fertilisation among multiple mates. When pre-copulatory inbreeding avoidance and polyandry were both fixed ($M_{a}=-10$ and $P_{a}=1$), $F_{a}$ allele values evolved to similarly negative means as $M_{a}$ allele values in Figure 4A,B (see Supporting Information Figure S12).

>>>>

After costs are realised, each surviving female chooses $N_{males}$ males to mate with, where $N_{males}$ is calculated by sampling from a Poisson distribution such that $N_{males} = Poisson(P_{p}) + 1$. This ensures that all surviving females choose at least one mate and generates each female's realised degree of polyandry with some stochastic variation around the expected mean $N_{males}$ of $P_{p}+1$.

@@

Most often, $N_{males}$ will be smaller than the total number of available males [@Duthie]. Each female then chooses her $N_{males}$ mates based on her pre-copulatory inbreeding avoidance phenotype ($M_{p}$). Negative or positive $M_{p}$ values cause a female to avoid or prefer mating with kin, respectively, whereas $M_{p}=0$ causes a female to mate randomly with respect to kinship.

@@

The probability that an offspring of female $i$ is sired by any one of her mates $j$ is calculated by assigning a perceived fertilisation quality to each $j$ with respect to $i$, $Q^{f}_{ij}$. Perceived fertilisation quality $Q^{f}_{ij}$ is calculated in the same way as perceived mate quality $Q^{m}_{ij}$, such that if female $i$ has a strategy of post-copulatory inbreeding avoidance ($F_{p}<0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = (-F_{p} \times k_{ij} + 1)^{-1}$. If the female has strategy of post-copulatory inbreeding preference ($F_{p}>0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = F_{p} \times k_{ij} + 1$. A relative quality ($q^{f}_{ij}$) is then calculated for each male by dividing his $Q^{f}_{ij}$ by the sum of the $Q^{f}_{ij}$ values across all of a female's mates. These $q^{f}_{ij}$ values, which lie between zero and one, define the probability of paternity. Females produce $n$ offspring, so a female $i$ samples from her mates $n$ times independently and with replacement with a probability of $q^{f}_{ij}$ for each male $j$ to determine the realised distribution of sires. Offspring have equal probability of being female or male. After offspring production, all female and male adults die so that generations are non-overlapping.

@@

The viability of a female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a log-linear function of her kinship with the sire $j$ of the offspring ($k_{ij}$) and inbreeding depression slope $\beta$,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
Here, $\beta$ models the number of haploid lethal equivalents that exist as deleterious recessive alleles in the gametes of $i$ and $j$, and which might be homozygous in offspring and reduce viability. Equation 1 assumes independent allelic effects, generating multiplicative effects on offspring viability [@Morton1956; @Mills1994]. It also assumes that inbreeding does not covary with inbreeding load (i.e., no purging). This formulation ensures that the relationship between $k_{ij}$ and the magnitude of inbreeding depression in offspring is consistent across replicate simulations and is justified because previous genetically-explicit modelling [@Duthie2016a] showed that inbreeding avoidance in biparental populations has a negligible effect on load given small-effect deleterious mutations [see also @Wang1999; @Guillaume2006]. We model inbreeding depression as having an absolute rather than relative effect on offspring viability (i.e., hard rather than soft selection), so that the effect of $\beta$ is consistent across generations and different parameter combinations. We assume that inbreeding always decreases offspring viability (i.e., $\beta > 0$, giving inbreeding depression but no outbreeding depression). Therefore, because $0 \leq k_{ij} \leq 1$, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero (if $-\beta \times k_{ij}$ is very negative) and one (if $-\beta \times k_{ij} = 0$). We therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability (versus mortality) using a Bernoulli trial. Offspring that are viable after the Bernoulli trial become adults. Given our current objectives, we simulate evolution under conditions where inbreeding avoidance is adaptive due to strong inbreeding depression, not where inbreeding preference is adaptive due to weak or zero inbreeding depression [@Parker1979; @Kokko2006; @Duthie2016a] or outbreeding depression [@Bateson1983; @Greeff2009]. However, as described above, positive $M_{p}$ and $F_{p}$ values resulting in inbreeding preference are not precluded from evolving, and could potentially arise due to mutation or drift.

@@

Simulations and analyses
----------------------------------------------

@@

To address how asymmetric costs associated with pre-copulatory and post-copulatory inbreeding avoidance and polyandry affect the long-term persistence of reproductive strategies, we quantify the change in $M_{a}$ and $F_{a}$ over generations in simulations where pre-copulatory inbreeding strategy was cost free ($c_{M}=0$) but post-copulatory inbreeding strategy was moderately costly ($c_{F}=0.02$), and vice versa. We compare evolutionary trajectories with those of a costly strategy in the absence of evolution of an alternative strategy (e.g., evolution of pre-copulatory inbreeding strategy when post-copulatory inbreeding strategy phenotype is fixed at zero, $F_{p}=0$). Previous modelling using similar genetic architecture suggests that a cost of $0.02$ imposes strong but not overwhelming direct negative selection on polyandry [@Duthie]. This value is therefore appropriate to illustrate the different evolutionary consequences that could result from asymmetrical costs (see Supporting Information).

To address how selection on an initially rare strategy of pre-copulatory or post-copulatory inbreeding avoidance, and resulting evolution, is affected by the other strategy of inbreeding avoidance already being fixed in the population, we first used exploratory simulations to quantify evolution of pre-copulatory inbreeding strategy, and of post-copulatory inbreeding strategy and associated polyandry, in isolation. Then, to test whether pre-copulatory inbreeding avoidance would evolve when adaptive polyandry and post-copulatory inbreeding avoidance were fixed, we initiated $M_{a}$ allele values at zero, but fixed $F_{a}$ allele values at $-10$ and $P_{a}$ allele values at $1$ (i.e., $F_{p}$ and $P_{p}$ were expressed but did not evolve further). Similarly, to test whether post-copulatory inbreeding avoidance would evolve when pre-copulatory inbreeding avoidance was already fixed, we initiated $F_{a}$ and $P_{a}$ allele values at zero but fixed $M_{a}$ allele values at $-10$. Consequently, because females have 10 diploid loci, when $M_{a}$ or $F_{a}$ alleles were fixed at $-10$, outbred females were $51$ times less likely to choose a full brother and $13.5$ times times less likely to choose a first cousin than a non-relative in pre-copulatory and post-copulatory choice, respectively.

In all simulations, we recorded mean values of $P_{a}$, $M_{a}$, and $F_{a}$ in each generation and present these values over generations to infer selection on phenotypes ($P_{p}$, $M_{p}$, and $F_{p}$). Each combination of parameter values simulated was replicated $40$ times, and grand mean values and standard errors of means are calculated in each generation across replicates. These analyses allowed us to infer how allele values changed over generations in response to costs, but also in response to the changing values of other alleles and therefore potential evolutionary feedbacks among reproductive strategies. We do not use statistical tests to interpret simulation results; such tests are inappropriate because simulations violate key assumptions of statistical hypothesis testing, and statistical power (and therefore p-values) is determined entirely by the number of simulation replicates [@White2014].

For all replicates, we set the maximum number of generations to 40000, which exploratory simulations and previous modelling [@Duthie] showed to be sufficient for inferring long-term dynamics of mean allele values and therefore selection on phenotypes. For all replicates, we set $\rho=5$ immigrants, which produced a range of kin and non-kin in each generation allowing females to express inbreeding strategies, and $n=8$ offspring, which was sufficient to keep populations consistently at carrying capacities and avoid population extinction. Values of $K_{f}$ and $K_{m}$ were both set to 100 because previous modelling showed that populations of this size are small enough that mate encounters between kin occur with sufficient frequency for selection on inbreeding strategy, but not so small that selection is typically overwhelmed by drift [@Duthie2016a].

@@

However, when both pre-copulatory and post-copulatory inbreeding avoidance could evolve, their relative evolutionary dynamics depended on their relative costs. When pre-copulatory but not post-copulatory inbreeding avoidance was costly ($c_{M}=0.02$ and $c_{F}=0$), pre-copulatory inbreeding avoidance initially evolved (i.e., $M_{a} < 0$) but then evolved back towards random mating (i.e., $M_{a} \approx 0$) following increasing evolution of post-copulatory inbreeding avoidance and polyandry (Figure 3B). Similarly, when post-copulatory but not pre-copulatory inbreeding avoidance was costly ($c_{F}=0.02$ and $c_{M}=0$), post-copulatory inbreeding avoidance initially evolved (i.e., $F_{a} < 0$) before evolving back to random fertilisation (i.e., $F_{a} \approx 0$) after ca $20000$ generations (Figure 3D).

Further simulations illustrate that such evolution of a high cost inbreeding strategy back toward random mating or random fertilisation in the presence of a relatively low cost alternative strategy is a general consequence of differential selection on each strategy induced by cost asymmetry within a range of relatively small costs, not specific to values of $c_{F}$ and $c_{M}$ of 0 and 0.02. For example, cost-specific evolution of pre-copulatory or post-copulatory inbreeding avoidance also occurred given non-zero small costs (e.g., $c_{F}$ and $c_{M}$ values of 0.01 and 0.03 and vice versa; Figure S9) and smaller cost asymmetries (e.g., $c_{F}$ and $c_{M}$ values of 0.01 and 0.02 and vice versa; Figure S10).

When pre-copulatory inbreeding avoidance was costly, allowing evolution of cost free post-copulatory inbreeding avoidance greatly facilitated evolution of polyandry (Figure 3A versus 3B). However when post-copulatory inbreeding avoidance was costly, allowing evolution of cost free pre-copulatory inbreeding avoidance caused $P_{a}$ alleles to decrease to very negative values, reducing expression of polyandry (Figure 3C versus 3D; polyandry was cost free in all these simulations). Results for all possible cost combinations of $0$ and $0.02$, including costly polyandry, are provided in Supporting Information (Figure S11).

@@

When polyandry alleles ($P_{a}$) were fixed to be positive so that all females were expected to mate multiply and post-copulatory inbreeding allele ($F_{a}$) values were fixed for adaptive inbreeding avoidance, pre-copulatory inbreeding avoidance evolved (i.e., $M_{a}$ values became increasingly negative; Figure 4A). Such evolution still occurred, but to a much smaller degree, when pre-copulatory inbreeding avoidance was costly (Figure 4B). However, after $40000$ generations, $M_{a}$ values were less negative when post-copulatory inbreeding avoidance and polyandry were fixed at non-zero values than when they also evolved from initial values of zero ($-15.45$ vs. $-17.62$; compare the blue lines in Figures 4A versus Figure 1B). This implies that pre-existence of fixed post-copulatory inbreeding avoidance can weaken selection and subsequent evolution of pre-copulatory inbreeding avoidance, but does not necessarily preclude it.

In contrast, when pre-copulatory inbreeding allele ($M_{a}$) values were fixed for adaptive inbreeding avoidance, mean $F_{a}$ allele values did not consistently become negative over generations (Figure 4C,D). This implies that existence of fixed pre-copulatory inbreeding avoidance can prevent evolution of post-copulatory inbreeding avoidance. In these simulations, mean $P_{a}$ allele values consistently decreased over generations, reflecting selection against polyandry regardless of whether or not post-copulatory inbreeding avoidance was costly (Figure 4C,D). Consequently, when $c_{F}=0$, $F_{a}$ allele values had no effect because females were almost exclusively monandrous, resulting in high drift of $F_{a}$ values (resulting in variation among replicates illustrated by the wide standard errors in Figure 4C). However, when $c_{F}=0.02$, $F_{a}$ values remained near zero to minimise direct costs. The lack of selection for post-copulatory inbreeding avoidance when pre-copulatory inbreeding avoidance was fixed was driven by a lack of polyandry, and therefore an inability of females to bias fertilisation among multiple mates. When pre-copulatory inbreeding avoidance and polyandry were both fixed ($M_{a}=-10$ and $P_{a}=1$), $F_{a}$ allele values evolved to similarly negative means as $M_{a}$ allele values in Figure 4A,B (see Supporting Information Figure S12).
